---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Victoria Thompson"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
getwd() #get working directory

#load packages
library(tidyverse)
library(agricolae)
library(ggplot2)
library(lubridate)
library(dplyr)


#load in raw data
NTL.raw <- read.csv(here("Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"),
                    stringsAsFactors = TRUE)
                  
#put dates in date format
NTL.raw$sampledate <- mdy(NTL.raw$sampledate)
                    
#2
mytheme <- theme_gray(base_size = 12) +
  theme(axis.text = element_text(color = "darkblue"),
        legend.position = "bottom",
         plot.title = element_text(face = "bold", size = 16, 
                                   color = "black", hjust = 1),
         legend.position = "bottom")

# set default 
theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperature recorded during July does not change with depth 
across all lakes.
Ha: The mean lake temperature recorded during July changes with depth across 
all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
NTL.filtered <- NTL.raw %>% 
  filter(month(sampledate) == 7)  %>% #month 7 = July
  select(lakename, year4, daynum, depth, temperature_C) %>%  
  # Select specific columns
  drop_na() #drop entire cases with na's

#5
NTL.temp.v.depth <- ggplot(NTL.filtered, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.5, color = "blue") + #transparent points to better 
  #show density
  geom_smooth(method = "lm", color = "darkgreen", se = FALSE) +
  scale_y_continuous(limits = c(0, 35)) +
  labs(
    title = "Temperature by Depth in July Across Lakes",
    x = "Depth (m)",
    y = "Temperature (°C)"
  ) +
  mytheme    

#print figure
NTL.temp.v.depth

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that as depth increases, temperature decreases.
The points themselves are slightly curved, almost in a exponential pattern. The 
points are more linear at lower depths, and flatten as depth increases. This 
could indicate that there may be a non-linear regression that fits the data 
better.

7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7

#perform the regression
temp.depth.regression <- 
  lm(NTL.filtered$temperature_C ~ 
       NTL.filtered$depth)

#display results
summary(temp.depth.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The p value of the model is <2e-16, and the R^2 value is 0.7387. With
such a small p-value, the likelihood that the relationship between variables is
purely chance is very low. This means that the relationship is statistically 
significant, and that changes in depth are very likely 
to influence lake temperature. The R^2 value of 0.7387 suggests that 
approximately 73.87% of the variability in lake temperature can be explained 
by changes in depth. This indicates a strong relationship, as depth explains 
a large portion of the temperature variation. There were 9726 degrees of 
freedom. To determine how much temperature is predicted to change for every 
1m change in depth, the slope of the line of best fit model (y=mx+b; slope is 
"m") is used. The coefficient of depth is -1.94621, meaning that for everying 
1-meter increase in depth, the model predicts a decrease in temperature by 
about 1.95 degrees C. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 

9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

library(corrplot)

#subset any possible explanatory variables 
NTL.subset <- NTL.raw %>% 
  filter(month(sampledate) == 7)  %>% #month 7 = July
  select(year4, daynum, depth, temperature_C) %>% 
  na.omit()

NTLcor <- cor(NTL.subset)
corrplot(NTLcor , upper= "ellipse")

#the plot indicates that depth and daynum may have a role in predicting 
#temperature

#10
#running a regression with both depth and irradiance
Temp.multi.regression <- lm(data = subset(NTL.raw, month(sampledate) == 7), 
                   temperature_C ~ depth + daynum)
summary(Temp.multi.regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC suggested that the variables of daynum and depth play a role 
in predicting temperature. This new model explains 74.1% of variance, as derived
from the new R^2 value. This means that the combination of these variables
explains more of the variation than just depth alone, since the regression with 
just depth could only explain 73.87% of variation in temperature. 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}

```

>Answer: 
